Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-09-10 paper/main.tex
Updates

\begin{abstract}
Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. Despite these advancements, the
application of these algorithms in practical domains remains limited, with
researchers often preferring to manually construct DAGs based on their expert
knowledge. Even when using these algorithms, researchers must rely on their
expertise to correct mistakes or orient edges before the models can be used for
further analysis. Given this reliance on manual model construction and
modification, we propose a method inspired by specification search in
Structural Equation Models (SEMs) to assist researchers in iteratively
modifying DAGs. Our method employs implied Conditional Independence (CI) based
model testing, combined with a novel measure of partial association for mixed
data, to rank violations of implied CIs of the model in the data. This ranking
helps users to prioritize pair of variables between which to add or remove
edges. Users can then decide to remove or add edges while determining the
correct orientation of the edge based on their expertise. Empirically, we
demonstrate that a greedy approach (i.e., fixing the worst violation first)
achieves performance comparable to automated algorithms, provided the user can
correctly determine the edge orientation in $ 1 $ out of $ 3 $ edge additions.
Additionally, we show that in the absence of an expert, a Large Language Model
(LLM) can potentially be used to determine edge orientations. 
\end{abstract}

@@

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific inquiries, as it reveals the underlying mechanisms of
observed phenomena and informs effective interventions or policy decisions.
\emph{Causal Discovery} methods aim to uncover these relationships among random
variables from observational data. This problem has been extensively explored
in both the Directed Acyclic Graphs (DAGs) and Structural Equation Models
(SEMs) literatures, each adopting a different approach.

>>>>

\begin{abstract}
Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. Despite these advancements, the
application of these algorithms in applied domains remains limited, with
researchers often preferring to manually construct DAGs based on their expert
knowledge. This preference stems due to some practical issues with the
automated algorithms such as they can make obvious mistakes, and they output
the Markov Equivalence Class (MEC). Due to these issues, when using these
algorithms, users typically have to rely on expert knowledge to fix the
mistakes and to orient the MEC to a DAG. Given this reliance on manual model
construction, we propose an iterative method to assist researchers in manually
constructing DAGs by combining implied Conditional Independence (CI) based
model testing and utilizing their expert knowledge. Our approach is inspired by
specification search in Structural Equation Models (SEMs), where we utilize a
novel measure of partial association for mixed data to rank violations of
implied Conditional Independences in the model. Using this ranking users can
prioritize the pair of variables which need to be corrected by either adding an
edge between them or removing the edge based on their expertise to specify the
correct orientation. Empirically, we demonstrate that a greedy approach (i.e.,
fixing the worst violation first) achieves performance comparable to automated
algorithms, provided the user can correctly determine the edge orientation in $
1 $ out of $ 3 $ edge additions. Additionally, we show that in the absence of
an expert, a Large Language Model (LLM) can potentially be used to determine
edge orientations. 
\end{abstract}

@@

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific inquiries, as it reveals the underlying mechanisms of
observed phenomena and informs effective interventions or policy decisions.
\emph{Causal Discovery} methods aim to uncover these relationships among random
variables from observational data. Methods for causal discovery have been
developed in both the Directed Acyclic Graphs (DAGs) and Structural Equation
Models (SEMs) literatures, each adopting a different approach.
